Computing properties of independent sets by generic programming tensor networks \thanks{\funding{...}}

### 2022-01-05 paper/paper.tex
section 4 revision

\section{Independence polynomial}
\subsection{Independence polynomial}\label{sec:indpoly}
The independence polynomial is an important graph polynomial that contains the counting information of independent sets. It is defined as
\begin{equation}\label{eq:idpdef}
I(G, x) = \sum_{k=0}^{\alpha(G)} a_k x^k,
\end{equation}
where $a_k$ is the number of independent sets of size $k$ in $G$. The total number of independent sets is thus equal to $I(G, 1)$.
To compute the independence polynomial of a graph $G = (V, E)$, we reinterpret this problem as a tensor network contraction problem.
We map a vertex $i\in V$ to a label $s_i \in \{0, 1\}$ of dimension $2$ in a tensor network, where we use $0$ ($1$) to denote a vertex absent (exists) in the set.
For each label $s_i$, we defined a parametrized rank-one vertex tensor $W(x_i)$ indexed by it as
\begin{equation}
    W(x_i) = \left(\begin{matrix}
        1 \\
        x_i
    \end{matrix}\right).
\end{equation}
We use subscripts to index tensor elements, e.g. $W(x_i)_0=1$ is the first element associated with $s_i=0$ and $W(x_i)_1=x_i$ is the second element associated with to $s_i=1$.
Similarly, on each edge $(u, v)$, we define a matrix $B$ indexed by $s_u$ and $s_v$ as
\begin{equation}
    \qquad \quad 
       B = \left(\begin{matrix}
        1  & 1\\
        1 & 0
    \end{matrix}\right). \label{eq:edgetensor}
\end{equation}
The corresponding tensor network contraction gives
\begin{equation}\label{eq:idp}
    P(G, \{x_1, x_{2}, \ldots,x_{|V|}\}) = \sum\limits_{s_1, s_2, \ldots, s_{|V|} = 0}^{1} \prod\limits_{i=1}^{|V|} W(x_i)_{s_i} \prod\limits_{(i,j) \in E(G)} B_{s_i s_j},
\end{equation}
where the summation runs over all vertex configurations $\{s_1, s_{2}, \ldots,s_{|V|}\}$ and accumulates the product of tensor elements to the output $P$ (see Example \ref{eg:tensorcontraction} for a concrete example).
The edge tensor element $B_{s_{i}=1, s_{j}=1} = 0$ encodes the independent set constraint, meaning vertex $i$ and $j$ cannot be both in the independent set if they are connected by an edge $(i,j)$.
In the special case of $x_i = x$, the contraction result directly corresponds to the independence polynomial.
The connection can be understood as follows: the product over vertex tensor elements produces a factor $x^k$, where $k=\sum_i s_i$ counts the set size,
and the product over edge tensor elements gives a factor $1$ for a configuration being in an independent set and $0$ otherwise.

%In the implementation of the algorithm, we benefit from recent advances in tensor network contraction,
%where researchers working on quantum circuit simulation evaluate the tensor network by pairwise contracting tensors in a heuristic order~\cite{Gray2021, Pan2021}.
To evaluate \Eq{eq:idp}, summing up the products directly is apparently computational inefficient.
The standard approach to evaluate a tensor network is to contract two tensors a time with a certain order utilizing the associativity and commutativity of tensor elements.
A good contraction order can reduce the time complexity significantly, at the cost of having a space overhead of $O(2^{\text{tw}(G)})$~\cite{Markov2008}.
The pairwise tensor contraction also makes it possible to utilize basic linear algebra subprograms (BLAS) functions to speed up the computation for certain tensor element types.

@@

    The contraction of this tensor network can be done in a pairwise order utilizing the associativity, additive commutativity and multiplicative commutativity of tensor elements:
    \begin{align*}
        &\sum_{s_a,s_b,s_c,s_d,s_e} W(x_a)_{s_a} W(x_b)_{s_b} W(x_c)_{s_c} W(x_d)_{s_d} W(x_e)_{s_e} B_{s_a s_b} B_{s_b s_d} B_{s_c s_d} B_{s_a s_c} B_{s_b s_c} B_{s_d s_e}\\
        =&\sum_{s_b,s_c}\left(\sum_{s_d}\left(\left(\left(\left(\sum_{s_e}B_{s_d s_e}W(x_e)_{s_e}\right) W(x_d)_{s_d}\right) \left(B_{s_bs_d} W(x_b)_{s_b}\right)\right) \left(B_{s_cs_d} W(x_c)_{s_c}\right)\right)\right.\\
        &\phantom{XXX}\left.\left(B_{s_bs_c}\left(\sum_{s_a}B_{s_as_b}\left(B_{s_as_c}W(x_a)_{s_a}\right)\right)\right)\right)\\
        =&1 + x_a + x_b + x_c + x_d + x_e + x_ax_d + x_ax_e + x_cx_e + x_bx_e\\
        =&1+5x+4x^2 \qquad \quad (x_{i} = x)
    \end{align*}
\end{example}

@@

Then we can treat the tensor elements as polynomials and evaluate the polynomial directly.
Let us create a polynomial type, and represent a polynomial $a_0 + a_1 x + \ldots + a_k x^k$ as a coefficient vector $(a_0, a_1, \ldots, a_k) \in \mathbb{R}^k$, so, e.g., $x$ is represented as $(0, 1)$.

@@

We can see these operations are standard addition and multiplication operations of polynomials, and the polynomial type forms a commutative ring. The tensors $W$ and $B$ can thus be written as 

@@

    W^{\rm poly} = \left(\begin{matrix}

@@

        B^{\rm poly} = \left(\begin{matrix}

@@

By contracting the tensor network with the polynomial type, we have contraction result the exact representation of the independence polynomial.
However, using polynomial numbers suffers from a space overhead proportional to $\alpha(G)$ because each polynomial requires a vector of such size to store the coefficients. 

@@

With this approach, we do not incur the linear overhead in space. However, because the independence polynomial coefficients can have a huge order-of-magnitude range, if we use floating point numbers in the computation, the round-off errors can be significant for the counting of large independent set sizes.

@@

@@

This is an iterative algorithm that iterates over a sequence of large prime numbers until convergence.

@@

Then we solve \Eq{eq:lineareq} using the gaussian elimination on ${\rm GF}(p)$ to find the coefficient modulo $p$, $A_p \equiv (a_0, a_1, \ldots, a_{\alpha(G)})\pmod p$.

@@

If this number does not change compared with the previous iteration, it indicates the convergence of result and the program terminates.

@@

        $A_p = (a_0, a_1, \ldots, a_{\alpha(G)}) \pmod p = {\rm gaussian\_elimination}(X, (y_0, y_1, \ldots, y_{\alpha{G}}) \pmod p) $\;

>>>>

\section{Independence polynomial} \label{sec:indpoly}
% \subsection{Independence polynomial}
The independence polynomial is an important graph polynomial that contains the counting information of independent sets. It is defined as
\begin{equation}\label{eq:idpdef}
I(G, x) = \sum_{k=0}^{\alpha(G)} a_k x^k,
\end{equation}
where $a_k$ is the number of independent sets of size $k$ in $G$. The total number of independent sets is thus equal to $I(G, 1)$.
To compute the independence polynomial of a graph $G = (V, E)$, we reinterpret this problem as a tensor network contraction problem.
We map a vertex $i\in V$ to a label $s_i \in \{0, 1\}$ of dimension $2$ in a tensor network, where we use $0$ ($1$) to denote a vertex is absent (present) in the set.
For each label $s_i$, we defined a parametrized rank-one vertex tensor $W(x_i)$ as
\begin{equation}
    W(x_i) = \left(\begin{matrix}
        1 \\
        x_i
    \end{matrix}\right).
\end{equation}
We use subscripts to index tensor elements, e.g.\ $W(x_i)_0=1$ is the first element associated with $s_i=0$ and $W(x_i)_1=x_i$ is the second element associated with $s_i=1$.
Similarly, on each edge $(u, v)$, we define a matrix $B$ indexed by $s_u$ and $s_v$ as
\begin{equation}
    \qquad \quad 
       B = \left(\begin{matrix}
        1  & 1\\
        1 & 0
    \end{matrix}\right). \label{eq:edgetensor}
\end{equation}
The corresponding tensor network contraction gives
\begin{equation}\label{eq:idp}
    P(G, \{x_1, x_{2}, \ldots,x_{|V|}\}) = \sum\limits_{s_1, s_2, \ldots, s_{|V|} = 0}^{1} \prod\limits_{i=1}^{|V|} W(x_i)_{s_i} \prod\limits_{(i,j) \in E(G)} B_{s_i s_j},
\end{equation}
where the summation runs over all vertex configurations $\{s_1, s_{2}, \ldots,s_{|V|}\}$ and accumulates the product of tensor elements to the output $P$ (see Example \ref{eg:tensorcontraction} for a concrete example).
The edge tensor element $B_{s_{i}=1, s_{j}=1} = 0$ encodes the independent set constraint, meaning vertex $i$ and $j$ cannot be both in the independent set if they are connected by an edge $(i,j)$.
In the special case of $x_i = x$, the contraction result directly corresponds to the independence polynomial.
The connection can be understood as follows: the product over vertex tensor elements produces a factor $x^k$, where $k=\sum_i s_i$ counts the set size,
and the product over edge tensor elements gives a factor $1$ for a configuration being in an independent set and $0$ otherwise. The summation counts the number of independent sets of size $k$.

%In the implementation of the algorithm, we benefit from recent advances in tensor network contraction,
%where researchers working on quantum circuit simulation evaluate the tensor network by pairwise contracting tensors in a heuristic order~\cite{Gray2021, Pan2021}.
To evaluate \Eq{eq:idp}, directly summing up the products is computationally inefficient.
The standard approach to evaluate a tensor network is to contract two tensors at a time with a certain order utilizing the associativity and commutativity of tensor elements.
A good contraction order can reduce the time complexity significantly, at the cost of having a space overhead of $O(2^{\text{tw}(G)})$~\cite{Markov2008}.
The pairwise tensor contraction also makes it possible to utilize basic linear algebra subprograms (BLAS) functions to speed up the computation for certain tensor element types.

@@

    The contraction of this tensor network can be done in a pairwise order utilizing the associativity, additive commutativity, and multiplicative commutativity of tensor elements:
    \begin{align*}
        &\sum_{s_a,s_b,s_c,s_d,s_e} W(x_a)_{s_a} W(x_b)_{s_b} W(x_c)_{s_c} W(x_d)_{s_d} W(x_e)_{s_e} B_{s_a s_b} B_{s_b s_d} B_{s_c s_d} B_{s_a s_c} B_{s_b s_c} B_{s_d s_e}\\
        =&\sum_{s_b,s_c}\left(\sum_{s_d}\left(\left(\left(\left(\sum_{s_e}B_{s_d s_e}W(x_e)_{s_e}\right) W(x_d)_{s_d}\right) \left(B_{s_bs_d} W(x_b)_{s_b}\right)\right) \left(B_{s_cs_d} W(x_c)_{s_c}\right)\right)\right.\\
        &\phantom{XXX}\left.\left(B_{s_bs_c}\left(\sum_{s_a}B_{s_as_b}\left(B_{s_as_c}W(x_a)_{s_a}\right)\right)\right)\right)\\
        =&1 + x_a + x_b + x_c + x_d + x_e + x_ax_d + x_ax_e + x_cx_e + x_bx_e\\
        =&1+5x+4x^2 \qquad \quad (x_{i} = x).
    \end{align*}
\end{example}

@@

%Then we can treat the tensor elements as polynomials and evaluate the polynomial directly.
Let us create a polynomial type and represent a polynomial $a_0 + a_1 x + \ldots + a_k x^k$ as a coefficient vector $a = (a_0, a_1, \ldots, a_k) \in \mathbb{R}^k$, so, e.g., $x$ is represented as $(0, 1)$.

@@

These operations are standard addition and multiplication operations of polynomials, and the polynomial type forms a commutative ring. The tensors $W$ and $B$ can thus be written as 

@@

    W^{\rm PN} = \left(\begin{matrix}

@@

        B^{\rm PN} = \left(\begin{matrix}

@@

By contracting the tensor network with the polynomial type, we have the exact representation of the independence polynomial.
However, using the polynomial type suffers a space overhead proportional to $\alpha(G)$ because each polynomial requires a vector of such size to store the coefficients. 

@@

With this approach, we do not incur the linear overhead in space. However, because the independence polynomial coefficients can have a huge order-of-magnitude range, if we use floating point numbers in the computation, the round-off errors can be significant for the counting of large-size independent sets.

@@

\eqname{GF$(p)$}

@@

The algorithm iterates over a sequence of large prime numbers until convergence.

@@

Then we solve \Eq{eq:lineareq} using Gaussian elimination on ${\rm GF}(p)$ to find the coefficient modulo $p$, $A_p \equiv (a_0, a_1, \ldots, a_{\alpha(G)})\pmod p$.

@@

If this number does not change compared with the previous iteration, it indicates the convergence of the result and the program terminates.

@@

        $A_p = (a_0, a_1, \ldots, a_{\alpha(G)}) \pmod p = {\rm gaussian\_elimination}(X, (y_0, y_1, \ldots, y_{\alpha(G)}) \pmod p) $\;
